`r params$doc_title`

### 2022-11-22 scripts/MSc.Rmd
boxplots overlay

@@

# try starting with n=50.000 -> selected 9 IVs
set.seed(1234)

## initialize
nsim <- 25
est_sim <- tibble(no_sim = 1:nsim)
#est_sim <- matrix(NA, 
#              nrow = nsim, 
#              ncol = 2*length(n)  # theta_sim and theta_se_sim
#              ) 
#colnames(est_sim) <- c(paste0("theta_sim_", n), paste0("theta_se_sim_", n))

runSimulation <- FALSE

## START SIM
if (runSimulation==TRUE) {
T0 = proc.time()[3]
for (i in 1:length(colnames_pval)) {
 threshold <- 5*10^-6
 se_update_BMI <- c()
 se_update_GBC <- c()

 numIV <- sum(my_data_harm[[colnames_pval[i]]] < threshold)

 if (numIV<5) { # start with 5 IVs only (sample size ~50,000)
   next
 }

 # FILTER DATA
 sim_dat <- my_data_harm %>% 
   filter(!!sym(colnames_pval[i]) < threshold) %>% 
   select(SNP, eaf.bmi, beta.bmi, p.value.bmi, eaf.gbc, beta.gbc, p.value.gbc)

 # get simulated se_{sample size}
 for (j in 1:nrow(sim_dat)) {
   # as done before in section "simulating on empirical data"
   se_update_BMI[j] <- abs(sim_dat[["beta.bmi"]][j]/-qnorm(p=(sim_dat[["p.value.bmi"]][i]/2),mean=0)*sqrt(n.orig)/sqrt(n[i]))
   se_update_GBC[j] <- abs(sim_dat[["beta.gbc"]][j]/-qnorm(p=(sim_dat[["p.value.gbc"]][i]/2),mean=0)*sqrt(n.orig)/sqrt(n[i]))
 }

 sim_dat <- sim_dat %>%
     bind_cols(tibble(se_update_BMI=se_update_BMI, se_update_GBC=se_update_GBC))

 # simulate beta values from normal_distribution # -----------------------------------------------------
 for (k in 1:nsim) {
   sim_dat <- sim_dat %>% 
     mutate(beta.bmi_sim_norm = purrr::map2_dbl(.x = beta.bmi, .y = se_update_BMI, .f = ~rnorm(n = 1, mean = .x, sd = .y))) %>% 
     mutate(beta.gbc_sim_norm = purrr::map2_dbl(.x = beta.gbc, .y = se_update_GBC, .f = ~rnorm(n = 1, mean = .x, sd = .y))) %>% 
     mutate(#explained_variance_X_sim = explained_variance_numeric(eaf = eaf.bmi, beta = beta.bmi), 
         explained_variance_X2_sim = explained_variance_numeric2(maf=eaf.bmi, 
                                                           beta=beta.bmi_sim_norm, 
                                                           se_beta=se_update_BMI, 
                                                           samplesize=n[i]
         ),
        # explained_variance_GSD = explained_variance_binary(PA = eaf.gsd, 
        #                                                    RR1 = exp(beta.gsd), 
        #                                                    RR2 = exp(beta.gsd)^2, 
        #                                                    K = 0.1  # assumed prevalence in population
        #                                                    )$Vg,
         explained_variance_Y_sim = explained_variance_binary(PA = eaf.gbc, 
                                                            RR1 = exp(beta.gbc_sim_norm), 
                                                            RR2 = exp(beta.gbc_sim_norm)^2, 
                                                            K =1.2/100000  # assumed prevalence in population
                                                            )$Vg
         )

   # PERFORM MR_analysis
   mr.obj = MendelianRandomization::mr_input(
         bx = sim_dat$beta.bmi_sim_norm, 
         bxse = sim_dat$se_update_BMI, 
         # outcome
         by = sim_dat$beta.gbc_sim_norm, 
         byse = sim_dat$se_update_GBC, 
         snps = sim_dat$SNP,
         exposure = "Body mass index",
         outcome = "Gallbladder Cancer"
       ) 
   # CONMIX (as reference)
   res = MendelianRandomization::mr_conmix(mr.obj)
   colname_theta <- paste0("theta_sim_", format(n[i], scientific=FALSE))
   colname_theta_se <- paste0("theta_se_sim_", format(n[i], scientific=FALSE))
   colname_pve_X <- paste0("pve_X_sim_", format(n[i], scientific=FALSE))
   colname_pve_Y <- paste0("pve_Y_sim_", format(n[i], scientific=FALSE))

   ## store results
   est_sim[k, colname_theta] <- res$Estimate
   CIlength = res$CIUpper-res$CILower
   if (length(CIlength)>1) print("conmix multimodal")
   est_sim[k, colname_theta_se] <- sum(CIlength)/1.96/2  ## Caution: this may be problematic 
   # add PVE cols
   est_sim[k, colname_pve_X] <- sum(sim_dat$explained_variance_X2_sim)
   est_sim[k, colname_pve_Y] <- sum(sim_dat$explained_variance_Y_sim)

   rm(res)
 }
 rm(se_update_BMI, se_update_GBC)
 print(paste0("|||-----------------------Run finished for sample size: ", format(n[i], scientific=FALSE), " -----------------------|||"))
}

T1 = proc.time()[3]
timediff = T1-T0

saveRDS(est_sim, paste0("../output/Rdata/", Sys.Date(), "_est_sim.rds"))
}

# read dat
est_sim <- readRDS(file = "../output/Rdata/2022-11-18_est_sim.rds")
```

@@

colname_theta <- colnames(est_sim)[str_detect(colnames(est_sim), pattern = "theta_sim")]
colname_theta_se <- colnames(est_sim)[str_detect(colnames(est_sim), pattern = "theta_se_sim")]

@@

ggplot(boxplot_dat, aes(x=forcats::fct_rev( as.factor(theta_sim_numeric)), y=value)) +
  geom_boxplot()+
  geom_jitter(width=0.1,alpha=0.2)+
  #scale_x_discrete(name = paste0("Size of sample ", as.roman(1)),  breaks = c(" 50000", "100000", "200000", "300000"), )+ 
  ylab(expression(paste(theta[sim], " with ", n[sim]== 25 ) )  ) +
  labs(title = "Estimates from beta simulation",
       caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
       )+
  coord_flip()+
  theme_few()
```

>>>>

We simulate

@@

## SIMULATION OUTSOURCED TO "./scripts/simstudy.R"

# read dat from "simstudy.R"
est_sim <- readRDS(file = "../output/Rdata/2022-11-18_est_sim.rds")

est_sim <- readRDS(file = "../output/Rdata/2022-11-22_est_sim_NSIM_5.rds")
```

@@

colname_theta <- colnames(est_sim)[str_detect(colnames(est_sim), pattern = "theta_sim")]
colname_theta_se <- colnames(est_sim)[str_detect(colnames(est_sim), pattern = "theta_se_sim")]
colname_pve_X <-  colnames(est_sim)[str_detect(colnames(est_sim), pattern = "pve_X_sim")]
colname_pve_Y <-  colnames(est_sim)[str_detect(colnames(est_sim), pattern = "pve_Y_sim")]

@@

ggplot(boxplot_dat, aes(x=forcats::fct_rev( as.factor(theta_sim_numeric)), y=value)) +
  geom_boxplot()+
  geom_jitter(width=0.1,alpha=0.2)+
  geom_hline(yintercept=  args$theta, col=ggthemes_data$few$colors$Dark[9,2][[1]], alpha=0.6, linetype=2, size=0.8)+
  annotate(geom = "text", x = "300000", y = args$theta+0.5,
           label= expression( paste("ref: ", theta[X] ))  , 
           color= ggthemes_data$few$colors$Dark[9,2][[1]]
             )+
  scale_x_discrete(name = paste0("Size of sample ", as.roman(1)),  breaks = c(" 50000", "100000", "150000", "200000", "250000", "300000"), )+ 
  ylab(expression(paste(theta[sim], " with ", n[sim]== 25 ) )  ) +
  labs(title = "Estimates from beta simulation",
       caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
       )+
  coord_flip()+
  theme_few()
```

```{r}
# EXPOSURE PVE
est_sim %>% 
  select(!!!(colname_pve_X)) %>% 
  pivot_longer(everything(), names_to = "PVE_X", values_to = "value_X") %>% 
  cbind(sample_size_filtered) %>% as_tibble() -> pve_X_sim

est_sim %>% 
  select(!!!(colname_pve_Y)) %>% 
  pivot_longer(everything(), names_to = "PVE_Y", values_to = "value_Y") %>% 
  cbind(sample_size_filtered) %>% as_tibble() -> pve_Y_sim

boxplot_overlay_dat <- pve_X_sim %>% 
  cbind(pve_Y_sim %>% select(-sample_size_filtered)) %>% 
  as_tibble()

fig01b.update +
  geom_boxplot(data=boxplot_overlay_dat, aes(x=sample_size_filtered, y=value_X*100, group=sample_size_filtered))

  geom_boxplot(data = test, aes(x=sample_size_filtered, y=value*100, # *100 to get percent values
                                group=sample_size_filtered), 
               col=ggthemes_data$few$colors$Dark[3,2][[1]],
               alpha=0.6)+

ggplot(test, aes(x=as.factor(sample_size_filtered), y=value))+
  geom_boxplot()

plot_data1b <- df_cum_expl_var %>% 
  pivot_longer(cols = c(cum_expl_varX2, cum_expl_varU, cum_expl_varY)) %>% 
  mutate(name = factor(name, labels = c("% var(U)", "% var(X)", "% var(Y)")),
         name = factor(name, levels = c("% var(X)", "% var(Y)", "% var(U)") )
         )
## plot
ggplot(plot_data1b, aes(x=n, y=value, col=name))+
  geom_point(alpha=0.6)+
  geom_line(alpha=0.6)+
  geom_line(aes(x=n, y=count_IVs_scaled, col="Number of IVs"), linetype="dotdash")+
  scale_x_continuous(name = paste0("Size of sample ", as.roman(1)),  # 2-sample-MR or label "Study Size"
                     labels = comma, 
                     breaks = N_annotation, #c( 10000, n[7:length(n)]),
                     limits = c(10000, 300000))+
  #ylab("% variance explained by IVs")+
  labs(title = "Figure 1",
       col='', 
       caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
       ) +
  scale_y_continuous(
    # Features of the first axis
    name = "% variance explained by IVs",
    # Add a second axis and specify its features
    sec.axis = sec_axis(~./scale_factor, name="Number of IVs"),
    limits = c(0,5)
  ) + 
  #geom_boxplot(data=boxplot_overlay_dat, aes(x=sample_size_filtered, y=value_X*100, group=sample_size_filtered))
  geom_boxplot(data = boxplot_overlay_dat, aes(x=sample_size_filtered, y=value_X*100, # *100 to get percent values
                                group=sample_size_filtered), 
               col=ggthemes_data$few$colors$Dark[3,2][[1]],
               alpha=0.2, size=0.2)+
  geom_boxplot(data = boxplot_overlay_dat, aes(x=sample_size_filtered, y=value_Y*100, # *100 to get percent values
                                group=sample_size_filtered), 
               col=ggthemes_data$few$colors$Dark[4,2][[1]],
               alpha=0.2, size=0.2)+
  theme_few()+
  #scale_colour_few() +
  #scale_color_manual(name = "", values = c("% var(U)" = "red", 
  #                                         "% var(X)" = "blue", 
  #                                         "% var(Y)" = "darkblue", 
  #                                         "Number of IVs" = "black"
  #                                         ))
  #
  scale_color_manual(name = "", values = c("% var(U)" = ggthemes_data$few$colors$Dark[2,2][[1]], 
                                           "% var(X)" = ggthemes_data$few$colors$Dark[3,2][[1]], 
                                           "% var(Y)" = ggthemes_data$few$colors$Dark[4,2][[1]], 
                                           "Number of IVs" = "black"),
                     # order
                     breaks = c( "Number of IVs", "% var(X)", "% var(Y)", "% var(U)")
  )
```
